Protein-intrinsic properties and context-dependent effects regulate pioneer-factor binding and function

### 2023-10-20 manuscript/manuscript_sections/_04_Discussion.qmd
update analysis and MS based on reviewer comments

To identify the mechanisms that regulate tissue-specific PF engagement with chromatin, we exogenously expressed Zld, Grh, and Twi in S2 cells where they are not normally expressed.
Leveraging prior data identifying binding sites for these factors *in vivo* and analyzing the wealth of data on chromatin structure in S2 cells, we identified features that govern PF activity.
We studied the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors.
Studying these three distinct TFs, which engage the genome through structurally distinct DBDs, allowed us to determine shared properties that regulate their pioneering activity.
Furthermore, our genome-wide analysis ofbinding, chromatin accessibility, and gene expression allowed us to separate PF occupancy and activity.
Collectively, we find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, protein concentration, DNA motif content, and protein domains outside the DBD (Fig. 6g).

Even when expressed at low levels Zld, Grh and Twi bind promiscuously to active, accessible regions with degenerate motifs or no detectable motifs at all.
When expressed at approximately physiological levels these factors all possess the capacity to bind to canonical motifs in closed chromatin.
However, these factors are largely excluded from repressed chromatin enriched for H3K9me3 or H3K27me3.
This binding to naïve chromatin is similar to what has previously been reported for other pioneer factors [@Soufi2012; @Donaghey2018].
Regions with high levels of H3K9me3 are resistant to binding by the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM).
Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions, suggesting that this repressive chromatin state is a barrier to PF binding [@Soufi2012].
Similarly, cell-type specific binding of the PF Pax7 is anti-correlated with levels of H3K27me3 [@Mayran2019].
Here, we specifically tested whether H3K27me3 regulates cell-type specific occupancy of Zld, Grh and Twi using the PRC2 inhibitor tazmetostat.
While H3K27me3 levels were decreased in our treatment, PF binding was largely unchanged.
Thus, unlike H3K9me3 in the case of OSKM, H3K27me3 is not a barrier to binding by the three factors assayed and suggest other features are restricting binding to these regions.

While depleting H3K27me3 levels did not result in Zld, Grh and Twi binding to additional regions, increased expression of both Zld and Grh resulted in binding to regions that were unbound at lower expression levels.
These results suggest that protein concentration and affinity for the target motif drives tissue-specific PF binding rather than the presence or absence of Polycomb-mediated silencing.
Tissue-specific PF occupancy is also likely regulated by the collection of cofactors expressed in a given tissue.
This may partially explain the presence of class VI sites that lack repressive histone modifications in S2 cells and are bound *in vivo* but not when expressed exogenously.
For example, in mammalian cell culture FOXA2 occupancy at a subset of tissue-specific binding sites was increased when GATA4 was present [@Donaghey2018].
Similarly, PARP-1 stabilizes Sox2 binding at subset of physiologically relevant sites in mouse embryonic stem cells [@Liu2017].
Unlike Zld and Grh, increased expression of Twi did not result in binding to regions bound in other tissues.
It is possible that additional cofactors, may be required for Twi occupancy at these regions.
Along with prior studies, our data suggest that PF occupancy is regulated by tissue-intrinsic features, including levels of PF expression, the complement of cofactors expressed and chromatin structure.

In addition to cell-type specific factors, we demonstrate that protein-intrinsic features control PF binding.
We show that regions outside the DBD of Zld and Grh are important for binding-site selection.
Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@lerner2022].
Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated a role for intrinsically disordered regions (IDRs) in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
For Zld, this disordered domain is required for transcriptional activation and promotes optimal nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
IDRs may contribute to TF binding through protein-protein interactions with other TFs, nonspecific interactions with DNA or histones, or by driving condensate formation [@Brodsky2021; @Staller2022].
Indeed, Zld is visualized in hubs within the nucleus, and these promote binding of additional transcription factors [@Mir2018; @Yamada2019].
Zld and Grh DBDs retain the capacity to bind to class I regions, despite the weaker motifs present at these sites.
Thus, the binding to class I regions is not due to interactions between IDRs and other proteins present at active chromatin.
Instead, this binding may be driven by the high affinity of the DBDs for nucleosome-free DNA.
To provide a more complete understanding of how cis-regulatory regions are established, we must further investigate how IDRs contribute to PF binding and chromatin accessibility.

The failure of the DBD alone to promote chromatin accessibility implicates a separation between the ability to bind nucleosomes and the capacity to promote chromatin accessibility.
While the DBD of Zld is sufficient to bind nucleosomes in vitro, regions outside the DBD promote this interaction [@McDaniel2019a].
Thus, nucleosome binding in vitro does not necessarily translate into in vivo occupancy of closed chromatin.
Local restructuring of nucleosome-DNA interactions may be insufficient for establishing accessible chromatin [@Michael2020; @Dodonova2020] and, instead, may stabilize PF binding to allow recruitment of cofactors, such as chromatin remodelers [@Judd2021; @King2017].
Furthermore, our identification of hundreds of class II regions, which are bound but not opened by exogenous PF expression, distinguishes chromatin binding and PF activity.
We propose that PF-mediated chromatin accessibility requires high chromatin occupancy.
This is supported by our data demonstrating a higher motif content at sites in which PF binding promotes accessibility as compared to sites which remain closed despite PF occupancy.
Furthermore, we identified a strong concentration dependence of Zld and Grh for chromatin opening.
This model would suggest that PF expression levels must be tightly controlled during development.
Indeed, misexpression of the PFs Grh and DUX4 leads to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel *cis*-regulatory elements and drive gene-regulatory networks to modulate cell fate.
A deeper understanding of how PF activity is regulated will have important implications for determining how diseases are caused by their misexpression and our ability to use these powerful factors for cellular reprogramming.

>>>>

To identify the mechanisms that regulate tissue-specific PF engagement with chromatin, we exogenously expressed Zld, Grh, and Twi in S2 cells where they are not normally expressed.
Leveraging prior data identifying binding sites for these factors *in vivo* and analyzing the wealth of data on chromatin structure and gene expression in S2 cells, we identified features that govern PF activity.
We studied the two well-defined PFs Zld and Grh, along with Twi, which possess features distinct from these two factors.
Studying these three distinct TFs, which engage the genome through structurally distinct DBDs, allowed us to determine shared properties that regulate their pioneering activity.
Furthermore, our genome-wide analysis of binding, accessibility, and gene expression allowed us to separate PF occupancy and activity.
Collectively, we find that pioneering activity requires a high level of chromatin occupancy that can be achieved through both protein-intrinsic and protein-extrinsic features such as local chromatin structure, cofactor expression, protein concentration, DNA motif content, and protein domains outside the DBD (Fig. 6g).

Even when expressed at low levels Zld, Grh and Twi bind promiscuously to active, accessible regions with degenerate motifs or no detectable motifs for the TF.
When expressed at approximately physiological levels these factors all possess the capacity to bind to canonical motifs in closed chromatin.
However, these factors are largely excluded from repressed chromatin enriched for H3K9me3 or H3K27me3.
This binding to naïve chromatin is similar to what has previously been reported for other pioneer factors [@Soufi2012; @Donaghey2018].
Regions with high levels of H3K9me3 are resistant to binding by the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM).
Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions, suggesting that this repressive chromatin state is a barrier to PF binding [@Soufi2012].
Similarly, cell-type specific binding of the PF Pax7 is anti-correlated with levels of H3K27me3 [@Mayran2019].
Here, we specifically tested whether H3K27me3 regulates cell-type specific occupancy of Zld, Grh and Twi using the PRC2 inhibitor tazmetostat.
While H3K27me3 levels were decreased in our treatment, PF binding was largely unchanged.
Thus, unlike H3K9me3 in the case of OSKM, H3K27me3 is not a barrier to binding by the three factors assayed and our analysis suggests other features, including the expression of additional TFs, are restricting binding to these regions.

While depleting H3K27me3 levels did not result in Zld, Grh and Twi binding to additional regions, increased expression of both Zld and Grh resulted in binding to regions that were unbound at lower expression levels.
These results suggest that protein concentration and affinity for the target motif drives tissue-specific PF binding rather than the presence or absence of Polycomb-mediated silencing.
This is consistent with recent work demonstrating the Zld-mediated chromatin opening in early embryos depends on motif affinity[@Brennan2023].
Tissue-specific PF occupancy is also likely regulated by the collection of cofactors expressed in a given tissue.
This may partially explain the presence of class VI sites that lack repressive histone modifications in S2 cells and are bound *in vivo* but not when expressed exogenously.
For example, in mammalian cell culture FOXA2 occupancy at a subset of tissue-specific binding sites was increased when GATA4 was present [@Donaghey2018].
Similarly, PARP-1 stabilizes Sox2 binding at subset of physiologically relevant sites in mouse embryonic stem cells [@Liu2017].
Unlike Zld and Grh, increased expression of Twi did not result in binding to regions bound in other tissues.
It is possible that additional cofactors may be particularly required for Twi occupancy at these regions, and it is clear that in the embryo Twi depends on Zld for binding.
However, our data also show that upon exogenous expression Twi possess some features of a PF.
This is similar to recent data demonstrating that TWIST1 is required for chromatin accessibility in human neural crest cells, and together support a model in which Twi possesses some features of a PF when expressed in certain tissues [@kim2023].
Along with prior studies, our data suggest that PF occupancy is regulated by tissue-intrinsic features, including levels of PF expression, the complement of cofactors expressed and chromatin structure.

In addition to cell-type specific factors, we demonstrate that protein-intrinsic features control PF binding.
We show that regions outside the DBD of Zld and Grh are important for binding-site selection.
Similarly, regions outside the DBDs of human PFs FOXA1 and SOX2 are required for robust binding of closed chromatin [@Lerner2023], and domains both N-terminal and C-terminal to the zinc-finger DBD of GAGA factor are required for stable genomic occupancy [@tang2022].
Increasing evidence suggests that eukaryotic transcription factors are not as modular as their bacterial counterparts and has demonstrated a role for intrinsically disordered regions (IDRs) in regulating chromatin occupancy[@Ferrie2022; @Brodsky2020] Both Zld and Grh contain long disordered domains.
For Zld, this disordered domain is required for transcriptional activation and promotes optimal nucleosome binding *in vitro* [@Hamm2015; @McDaniel2019a].
IDRs may contribute to TF binding through protein-protein interactions with other TFs, nonspecific interactions with DNA or histones, or by driving condensate formation [@Brodsky2021; @Staller2022].
Indeed, Zld is visualized in hubs within the nucleus, and these promote binding of additional transcription factors [@Mir2018; @Yamada2019].
Zld and Grh DBDs retain the capacity to bind to class I regions, despite the weaker motifs present at these sites.
Thus, the binding to class I regions is not due to interactions between IDRs and other proteins present at active chromatin.
Instead, this binding may be driven by the high affinity of the DBDs for nucleosome-free DNA.
To provide a more complete understanding of how cis-regulatory regions are established, we must further investigate how IDRs contribute to PF binding and chromatin accessibility.

The failure of the DBD alone to promote chromatin accessibility implicates a separation between the ability to bind nucleosomes and the capacity to promote chromatin accessibility.
While the DBD of Zld is sufficient to bind nucleosomes in vitro, regions outside the DBD promote this interaction [@McDaniel2019a].
Thus, nucleosome binding in vitro does not necessarily translate into in vivo occupancy of closed chromatin.
Local restructuring of nucleosome-DNA interactions may be insufficient for establishing accessible chromatin [@Michael2020; @Dodonova2020] and, instead, may stabilize PF binding to allow recruitment of cofactors, such as chromatin remodelers [@Judd2021; @King2017].
Furthermore, our identification of hundreds of class II regions, which are bound but not opened by exogenous PF expression, distinguishes chromatin binding and PF activity.
We propose that PF-mediated chromatin accessibility requires high chromatin occupancy.
This is supported by our data demonstrating a higher motif content at sites in which PF binding promotes accessibility as compared to sites which remain closed despite PF occupancy.
Furthermore, we identified a strong concentration dependence of Zld and Grh for chromatin opening.
This model would suggest that PF expression levels must be tightly controlled during development.
Indeed, misexpression of the PFs GRHL2 and DUX4 leads to diseases such as epithelial cancer or facioscapulohumeral muscular dystrophy (FSHD), respectively [@Campbell2018; @Reese2019].
Together our data identify protein intrinsic and extrinsic features that govern PF binding and activity, providing insights into how these factors define novel *cis*-regulatory elements and drive gene-regulatory networks to modulate cell fate.
A deeper understanding of how PF activity is regulated will have important implications for determining how diseases are caused by their misexpression and our ability to use these powerful factors for cellular reprogramming.

### 2023-10-20 manuscript/manuscript_sections/_05_Methods.qmd
update analysis and MS based on reviewer comments

Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
Expression of protein at physiological levels was achieved with 1000, 100, or 40 µM CuSO~4~ for Zld, Grh or Twi, respectively (Extended Data Fig. 1c-d; Extended Data Fig. 4a).
For experiments with sub- or supraphysiological protein levels (Fig. 5; Extended Data Fig. 9a-c), the following CuSO~4~ concentrations were used: 500 µM (Zld subphysiological), 1000 µM (Zld physiological) or 1500 µM (Zld supraphysiological); 25 µM (Grh subphysiological), 100 µM (Grh physiological) or 400 µM (Grh supraphysiological); 10 µM (Twi subphysiological), 40 µM (Twi physiological) or 160 µM (Twi supraphysiological).
For all induction experiments, cells were plated at 1×10^6^ cells per mL.
Unless otherwise specified, experiments were performed with cells harvested 48 hours after induction.

@@

Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: `-k 2 --very-sensitive --no-mixed --no-discordant -X 5000`.
Aligned reads were filtered to include only reads with a mapping quality score \> 30.
Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
Peak calling was performed using MACS2 version 2.2 with parameters `-g dm --call-summits`.
Data from ChIP experiments performed with the same antibody in wild-type cells were used as a control for peak calling.
Only peaks that were detected in all replicates were considered in downstream analysis.
bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters `–binSize 10`.
bigWig files were z-score normalized by subtracting the genome-wide average of all bins from each bin value and dividing by the standard deviation of all bins.

@@

@@

CUT&RUN was performed using EpiCypher reagents according to the manufacturer protocol.
2×10^5^ cells were used for each CUT&RUN reaction.
Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
0.5 µL anti-H3K27me3 (Cell signaling technologies), anti-Zld, anti-Grh, or rabbit IgG were used for CUT&RUN.
For H3K27me3 CUT&RUN, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction prior to addition of antibody.
Libraries were prepared using the NEBNext Ultra II kit (NEB).
During library preparation, cleanup steps were performed using a 1.1X ratio of Axygen paramagnetic beads, as recommended by EpiCypher.
PCR amplification was performed with the following conditions: 98°C for 45 seconds, then 14 cycles of 98°C for 15 seconds, 60°C for 10 seconds, and a final extension of 72°C for 1 minute.
Sequencing was performed on the Illumina NovaSeq 6000 with 150bp paired-end reads.

@@

@@

Previously published datasets analyzed in this study can be found in supplementary table 2.
All previously published data were analyzed in parallel with data from this study using analysis parameters described above.

@@

## Acknowledgements
We thank Alex Theis, Meghan Freund, and Andrew Mehle for helpful discussions and advice. We thank Julia Zeitlinger for generously sharing the Twist antibody. We acknowledge the University of Wisconsin-Madison Biotechnology Center and the NUSeq Core Facility for sequencing. TJG was supported by the National Institutes of Health (NIH) National Research Service Award T32 GM007215. Experiments were supported by a R35 GM136298 from NIH to MMH. MMH was also supported by a Vallee Scholar Award.

>>>>

Transcription factor expression was induced by adding CuSO~4~ to the cell culture media.
Expression of protein at approximately physiological levels was achieved with 1000, 100, or 40 µM CuSO4 for Zld, Grh or Twi, respectively (Extended Data Fig. 1c-d; Extended Data Fig. 4a).
For experiments with multiple protein levels (Fig. 5; Extended Data Fig. 9a-c), the following CuSO~4~ concentrations were used: 500 µM (Zld low), 1000 µM (Zld medium) or 1500 µM (Zld high); 25 µM (Grh low), 100 µM (Grh medium) or 400 µM (Grh high); 10 µM (Twi low), 40 µM (Twi medium) or 160 µM (Twi high); 100 µM (Zld DNA-binding domain low); 400 µM (Zld DNA-binding domain high); 20 µM (Grh DNA-binding domain low); 100 µM (Grh DNA-binding domain high).
The Zld or Grh DNA-binding domains were induced with 100 µM, or 20 µM CuSO~4~, respectively for immunofluorescence experiments (Extended Data Fig. 10).
For all induction experiments, cells were plated at 1×10^6^ cells per mL.
Unless otherwise specified, experiments were performed with cells harvested 48 hours after induction.

@@

Bowtie2 version 2.4.4 [@Langmead2012] was used to align ChIP-seq reads to the *Drosophila melanogaster* genome (version dm6) with the following non-default parameters: `-k 2 --very-sensitive --no-mixed --no-discordant -X 5000`.
Aligned reads were filtered to include only reads with a mapping quality score \> 30.
Reads mapping to unplaced scaffolds or the mitochondrial genome were discarded.
Peak calling was performed using MACS2 version 2.2 with parameters `-g dm --call-summits`.
Data from ChIP experiments performed with the same antibody in wild-type cells were used as a control for peak calling.
Thus, only peaks with statistically significantly enrichment (MACS2 q-value \< 0.5) in induced cells relative to wild-type cells are called as peaks.
Only peaks that were detected in all replicates were considered in downstream analysis.
bigWig files were generated using deepTools bamCoverage version 3.5.1 with parameters `–binSize 10`.
bigWig files were z-score normalized by subtracting the genome-wide average of all bins from each bin value and dividing by the standard deviation of all bins.

@@

## Drosophila strains and collection of brains for CUT&RUN

Drosophila strains and genetic crosses for the isolation of type II neuroblasts (neural stem cells) and for the ectopic expression of Zld in neural stem cells were described previously [@Larson2021a].
Brains from *brat^11/DF^* mutant larvae were used to perform CUT&RUN for endogenous Zld in neural stem cells.
For ectopic expression of Zld, brains were collected from *brat^11/DF,Tub-Gal80ts;UAS-HA-Zld^* larvae, in which *Worniu-Gal4* drives overexpression of HA-Zld in the neural stem cells.
Larvae were grown at 18°C until reaching the L3 stage, then shifted to 33°C for 24 hours to induce Zld overexpression.
50 brains were dissected into Schneider's medium.
Brains were centrifuged at 600 × *g* for 3 minutes and transferred to CUT&RUN wash buffer supplemented with 0.1% BSA.
Brains were dounced 10 times in a 1 mL dounce, centrifuged at 600 × *g* for 3 minutes and resuspended in 100 µL CUT&RUN wash buffer.

@@

CUT&RUN was performed using EpiCypher reagents according to the manufacturer protocol.
2×10^5^ cells, or material from 50 brains prepared as described above, were used for each CUT&RUN reaction.
Overnight incubation with antibodies was performed in antibody buffer containing 0.05% digitonin.
0.5 µL anti-H3K27me3 (Cell signaling technologies), anti-Zld, anti-Grh, or rabbit IgG were used for CUT&RUN.
For H3K27me3 CUT&RUN, a 1:5 dilution of K-MetStat Panel spike-in nucleosomes (EpiCypher, cat. \# 19-1002) was added to each reaction prior to addition of antibody.
Libraries were prepared using the NEBNext Ultra II kit (NEB).
During library preparation, cleanup steps were performed using a 1.1X ratio of Axygen paramagnetic beads, as recommended by EpiCypher.
PCR amplification was performed with the following conditions: 98°C for 45 seconds, then 14 cycles of 98°C for 15 seconds, 60°C for 10 seconds, and a final extension of 72°C for 1 minute.
Sequencing was performed on the Illumina NovaSeq 6000 with 150bp paired-end reads.

@@

## Motif analysis

Motif analysis was performed using the MEME suite[@Bailey2015], the memes R package[@Nystrom2021], and the universalmotif R package[@universalmotif].
To facilitate analysis of canonical Zld, Grh, or Twi motifs, position-weight matrices were derived from ChIP-seq data for each factor in the *Drosophila* embryos.
FIMO[@Grant2011] was used to find instances of the canonical motifs within ChIP-seq peaks for S2 cells.
Motifs with an adjusted p-value \< 1×10^-3^ were considered for analysis of motif number and score.
AME[@mcleay2010] was used to test for enrichment of known motifs from the FlyFactorSurvey database[@zhu2010].
Previously published RNA-seq data was used to pre-filter motifs to include only those for which the associate factor is expressed (RPKM \> 1) in at least one of the tissues being analyzed (embryo, neural stem cells, and wing discs).
Motifs enrichment was visualized using the normalized motif rank implemented by the memes package.

@@

Previously published datasets analyzed in this study can be found in Supplementary Table 2.
All previously published data were analyzed in parallel with data from this study using analysis parameters described above.

@@

## Acknowledgements

We thank Alex Theis, Meghan Freund, and Andrew Mehle for helpful discussions and advice.
We thank Julia Zeitlinger for generously sharing the Twist antibody.
We particularly thank Elizabeth Larson for sharing the neural stem cell data for HA-Zld.
We acknowledge the University of Wisconsin-Madison Biotechnology Center and the NUSeq Core Facility for sequencing.
TJG was supported by the National Institutes of Health (NIH) National Research Service Award T32 GM007215.
Experiments were supported by a R35 GM136298 and R01 NS111647 from NIH to MMH.
MMH was also supported by a Vallee Scholar Award.
MMH is a Romnes Faculty Fellow and Vilas Faculty Mid-Career Investigator.
